The Behavior Insight Design Framework: A Theoretical Integration for Advancing Applied Psychology in Interactive Systems

### 2025-07-03 manuscripts/theory/theory.qmd
Adds BID framework image and updates authorship
- Adds an image depicting the Behavior Insight Design (BID) framework to the manuscript.
- Removes a placeholder image.
- Removes a co-author, their affiliation, and a corresponding footnote per request; updates the manuscript accordingly.

---
title: "The Behavior Insight Design Framework: A Theoretical Integration for Advancing Applied Psychology in Interactive Systems"
subtitle: Behavioral Insight Design
author:
  - name: Jeremy R. Winget, PhD
    email: contact@jrwinget.com
    affiliations: 
        - id: independent
          name: Independent Researcher
          city: Chicago
          state: IL
          postal-code: 60660
    attributes:
        corresponding: true
  - name: Milena Eickhoff
    email: m.eickhoff@skimgroup.com
    affiliations:
        - id: skim
          name: SKIM
          department: Development
          address: 'Weena 695, B6.036'
          city: Rotterdam
          country: The Netherlands
          postal-code: 3013 AM
    note: |
      M.E. is employed by SKIM, a market research company. J.R.W. and M.E. were colleagues at SKIM during the initial development of the BID framework, but the theory and research presented was conducted independently and does not represent SKIM's official positions.
abstract: |
  Applied psychology faces a fundamental challenge in translating laboratory-derived principles to complex, real-world contexts where multiple psychological mechanisms operate simultaneously. The Behavior Insight Design (BID) framework addresses this theoretical gap through systematic integration of established psychological principles, advancing our understanding of how cognitive mechanisms interact in naturalistic environments. The framework operationalizes psychological theory through five sequential stages (Notice, Interpret, Structure, Anticipate, Validate) that map to fundamental phases of human information processing while enabling empirical investigation of mechanism interactions previously inaccessible through traditional isolated studies. BID advances psychological science in three critical ways: providing a structured platform for testing psychological predictions in ecologically valid environments, revealing novel interactions between cognitive load and bias susceptibility that emerge only in complex decision contexts, and demonstrating how established principles operate when combined in realistic environments. The framework contributes methodologically by enabling systematic investigation of multi-mechanism interactions, longitudinal process dynamics, and individual differences in naturalistic contexts while maintaining experimental rigor. Hypothetical study designs illustrate how BID enables investigation of cognitive load-bias interactions, expertise-dependent organizational effects, and individual-group optimization trade-offs that advance theoretical understanding beyond conventional experimental approaches. Rather than simply applying psychology to design, BID generates new theoretical insights about psychological processes while providing methodological innovations for conducting naturalistic psychological research that simultaneously advances basic science and practical application.
keywords: 
  - cognitive load
  - dual-process theory
  - decision making
  - bias susceptibility
  - individual differences
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Applied Psychology
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

@@

>>>>

---
title: "The Behavior Insight Design Framework: A Theoretical Integration for Advancing Applied Psychology in Interactive Systems"
author:
  - name: Jeremy R. Winget, PhD
    email: contact@jrwinget.com
    affiliations: 
        - id: independent
          name: Independent Researcher
          city: Chicago
          state: IL
          postal-code: 60660
    attributes:
        corresponding: true
abstract: |
  Applied psychology faces a fundamental challenge in translating laboratory-derived principles to complex, real-world contexts where multiple psychological mechanisms operate simultaneously. The Behavior Insight Design (BID) framework addresses this theoretical gap through systematic integration of established psychological principles, advancing our understanding of how cognitive mechanisms interact in naturalistic environments. The framework operationalizes psychological theory through five sequential stages (Notice, Interpret, Structure, Anticipate, Validate) that map to fundamental phases of human information processing while enabling empirical investigation of mechanism interactions previously inaccessible through traditional isolated studies. BID advances psychological science in three critical ways: providing a structured platform for testing psychological predictions in ecologically valid environments, revealing novel interactions between cognitive load and bias susceptibility that emerge only in complex decision contexts, and demonstrating how established principles operate when combined in realistic environments. The framework contributes methodologically by enabling systematic investigation of multi-mechanism interactions, longitudinal process dynamics, and individual differences in naturalistic contexts while maintaining experimental rigor. Hypothetical study designs illustrate how BID enables investigation of cognitive load-bias interactions, expertise-dependent organizational effects, and individual-group optimization trade-offs that advance theoretical understanding beyond conventional experimental approaches. Rather than simply applying psychology to design, BID generates new theoretical insights about psychological processes while providing methodological innovations for conducting naturalistic psychological research that simultaneously advances basic science and practical application.
keywords: 
  - cognitive load
  - dual-process theory
  - decision making
  - bias susceptibility
  - individual differences
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Applied Psychology
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

@@

![The Behavior Insight Design Framework](./bid-framework.png)

### 2025-07-03 manuscripts/theory/theory.tex
Adds BID framework image and updates authorship
- Adds an image depicting the Behavior Insight Design (BID) framework to the manuscript.
- Removes a placeholder image.
- Removes a co-author, their affiliation, and a corresponding footnote per request; updates the manuscript accordingly.

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Applied Psychology}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Behavior Insight Design Framework: A Theoretical Integration for Advancing Applied Psychology in Interactive Systems},
  pdfauthor={Jeremy R. Winget, PhD; Milena Eickhoff},
  pdfkeywords={cognitive load, dual-process theory, decision
making, bias susceptibility, individual differences},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The Behavior Insight Design Framework: A Theoretical Integration
for Advancing Applied Psychology in Interactive
Systems \\\large{Behavioral Insight Design} }
\author[1]{Jeremy R. Winget, PhD%
\corref{cor1}%
}
 \ead{contact@jrwinget.com} 
\author[2]{Milena Eickhoff%
%
\fnref{fn1}}
 \ead{m.eickhoff@skimgroup.com}

\affiliation[1]{organization={Independent
Researcher},city={Chicago},postcode={60660},postcodesep={}}
\affiliation[2]{organization={SKIM, Development},addressline={Weena 695,
B6.036},city={Rotterdam},country={The
Netherlands},countrysep={,},postcode={3013 AM},postcodesep={}}

@@

\fntext[fn1]{M.E. is employed by SKIM, a market research company. J.R.W.
and M.E. were colleagues at SKIM during the initial development of the
BID framework, but the theory and research presented was conducted
independently and does not represent SKIM's official positions.}

@@

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Applied Psychology}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Behavior Insight Design Framework: A Theoretical Integration for Advancing Applied Psychology in Interactive Systems},
  pdfauthor={Jeremy R. Winget, PhD},
  pdfkeywords={cognitive load, dual-process theory, decision
making, bias susceptibility, individual differences},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The Behavior Insight Design Framework: A Theoretical Integration
for Advancing Applied Psychology in Interactive Systems}
\author[1]{Jeremy R. Winget, PhD%
\corref{cor1}%
}
 \ead{contact@jrwinget.com}

\affiliation[1]{organization={Independent
Researcher},city={Chicago},postcode={60660},postcodesep={}}

@@

@@

\begin{figure}[H]

{\centering \pandocbounded{\includegraphics[keepaspectratio]{./bid-framework.png}}

}

\caption{The Behavior Insight Design Framework}

\end{figure}%
